---
title: "Preliminary Analysis Party Frames Migration"
author: "Nicolai Berk"
format: html
editor: visual
---

## Introduction

```{r setup}

knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = T, fig.height=6, fig.width=10, fig.align='center', eval = T)

# Setup ####
library(tidyverse)
library(here)
library(data.table)
library(word2vec)


# load self-made functions
for (filename in list.files(here("code/functions"))){
  source(here(paste0('code/functions/', filename)))
}

```

This document contains some preliminary analyses of frames employed in news coverage and party communication about migration. All analyses are based on a subsample and supposed to inform any analyses on a broader sample later on.

## Explore Embeddings

We start by exploring the quality of our word embeddings:

```{r wordembeddings}

# load model
w2v_model <- 
  read.word2vec(here("data/processed/embeddings/w2v_news_sample.bin"))

# read vectors
load(here("data/processed/embeddings/w2v_news_vectors_sample.Rdata"))

vector <- emb["Ausländer", ]
predict(w2v_model, vector, type = "nearest", top_n = 10)

  

```

## Frames in Migration coverage

## Frames in Party Coverage

## Association of Party and Migration Coverage
